JD notes on effects

### 2022-07-09 manuscript/body.tex
x-axis labels

We now illustrate the construction of outcome plots, compare with other packages and also demonstrate that the mean-based and whole-sample-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias. In addition, we demonstrate that whole-sample-based approach can be used for bias correction.

@@

The \proglang{R} software package \pkg{margins} \cite{lenth2018package} generates marginal effects. To illustrate the distinction between effects and marginal effects, we use the simulated data described in \nameref{S1_Appendix}. 
%
\begin{figure}
\centering
\includegraphics{cubic_varpred_margins_pred.ggp.pdf}
\caption{{\bf Effect and marginal effect plots.} \NEW{A:  The effect plot. B: The marginal effect (ME) at representative values of the focal predictor \code{age} together with the $95\%$ confidence intervals. For simple linear models with no interactions, the ME should be constant across all the values of the focal predictor; and it would correspond to the slope of the central estimate of central estimate. However, in this example, the focal predictor is a cubic polynomial, and so, the effect is not constant but a function of the partial derivatives of the corresponding model variables.}}
\label{fig:qoi_age_pred_plot}
\end{figure}
%

@@

To illustrate the distinction between the two approaches, we simulate a multivariate model described in \nameref{S1_Appendix}. We also use the same model to illustrate how predictions and effects differ in the presence of higher order interactions in focal or non-focal input variables. We then generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal input variable (a cubic polynomial focal input variable); and 2) one with \code{Wealthindex} as the focal in put variable (linear focal input variable). In both cases the, the center point was used as the anchor for the effects, as shown in \FREF{pred_cubic_plots}.

\begin{figure}
\begin{center}
\includegraphics{cubic_predictors_preds.ggp.pdf}
\end{center}
\caption{{\bf Prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal input variable is a cubic polynomial. B: The focal input variable is linear, with polynomial non-focal input variable. The central curve and trend lines are the same for predictions and effects plots in each case. However, the confidence intervals for the effects are narrower than those of the predictions in both cases. The default anchor for generating effects is the center point. For simple OLS models (or simple focal input variables), the center point is a point within the range of focal values; and therefore the confidence interval lines intersect at that that point -- B. In the case of higher order interactions such as cubic polynomials, the center point is not a point but a combination of all the predictor variables associated with the cubic polynomial, consequently, curves will not necessarily cross at a point -- A. The horizontal black and red dashed lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively.}
\label{fig:pred_cubic_plots}
\end{figure}

Predictions are the standard curves (wider); they tell us what we expect to see if we know the value of the focal input variable. In multivariate models, at times, we may want to focus on the effect of the focal input variable only. Unlike prediction plot, which captures the uncertainty in the entire model, effect plot focuses on the uncertainty of the focal input variable only. In this case, effects styled plots which are usually narrower around the anchor are preferable over the conventional prediction plots which generally focus on the \emph{total} effects.

In the absence of complex higher order interactions or transformations on the focal input variable, in simple OLS models we would expect the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and intersect with the trend lines at the center point, for the effects plot. On the other hand, if the focal input variable is characterized by complex interactions or any other form of transformations, the center point is not necessarily a point in the range of focal input values. In this case, the effects will be narrower than the predictions plots but will not necessarily intersect.

@@

Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal input variable(s), random effects or complex interactions. To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that age has a very small effect size in comparison to wealth index, and compared their effects on the predicted probability of improved water quality as shown in \FREF{pred_bin_plots}.

\begin{figure}
\begin{center}
\includegraphics{glm_two_predictor_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and whole-sample-based predictions.} The \emph{truth} is the prediction based on the true simulation parameters. If the focal input variable has small effect size, mean-based and whole-sample-based approaches produce different predictions -- A. In the case of strong effect size, however, the two approaches produce similar predictions -- B. In both cases, the whole-sample-based predictions are close to the truth. The difference in mean-based and whole-sample-based is due to the bias induced by the non-focal variables and the nonlinear link function in the logistic model. Mean-based approach suffers from nonlinear averaging; and if focal input variable has small effect size, the bias is even pronounced since the effect of nonlinear averaging is driven by the non-focal input variable with the strong effect. However, this is not the case if the non-focal input variables have small effect since the bias they induce is equally likely to be small. Since whole-sample-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the nonlinear averaging for the reasons already mentioned. The horizontal dashed lines correspond to the respective averages. The vertical dashed lines represent the center point, and the point at which they intersect the horizontal lines represents the expected ``perfect'' prediction at the model center. The grey points are binned observations -- observed proportions of improved water quality in each bin.} 
\label{fig:pred_bin_plots}
\end{figure}

@@

In \FREF{pred_bin_prediction_effects_plots} below, we zoom into \FREF{pred_bin_plots}A and compare the prediction and effect styled plots.

\begin{figure}
\begin{center}
\includegraphics{glm_two_predictor_preds_effects.ggp.pdf}
\end{center}
\caption{{\bf Prediction and effect styled plots and bias correction.} The central curves for predictions and effects are the same in the respective case. These patterns are  consistent with the ones in \FREF{pred_bin_plots}A with respect to the \emph{truth}. Further, the prediction and effect CIs exhibit patterns similar to the ones we previously described and generalizes in a similar way to the ones in simple linear models.
}
\label{fig:pred_bin_prediction_effects_plots}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics{mediate_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and whole-sample-based predictions for mediated effects for the mediated (multivariate) and non-mediated (univariate) models.} A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictions are identical since there is no any other additional (non-focal input variables) sources of bias. Consequently, for both approaches, the average predicted probability is very close to the observed proportion. B: When the mediator variable is included,  $x$ has no direct effect on $z$ and as a result the predictions do not align with the observations. However, the whole-sample-based approach still closely approximates the average proportion in the simulated data. The horizontal dashed lines are the respective average predictions and observed proportions. The vertical dashed black represents the center point, and the point at which it crosses the horizontal lines represents the expected ``perfect'' prediction at the model center. The grey points are the binned observations.}
\label{fig:pred_mediated_plots}
\end{figure}

From \FREF{pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$. This is the total effect of $x$, which only tells us the influence of $x$ on $z$. By ignoring $y$ in the model, we are still able to capture the effect of $x$ and closely match the observed values using both approaches. However, if there were additional non-focal predictors, we would expect to see the differences similar to those in \FREF{pred_bin_plots}A. Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $z$ and as a result, our predictions do not necessarily match the observed binned observations (see \FREF{pred_mediated_plots}B).

@@

The argument and results we present in this paper support a greater need for a shift on focus on how to present predictions from generalized models. For example, the effect-styled confidence intervals could provide more clarity concerning the uncertainty due to the input variable of interests as opposed to the conventional way of incorporating everything.

>>>>

We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots and also demonstrate that the mean-based and whole-sample-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias. In addition, we demonstrate that whole-sample-based approach can be used for bias correction.

@@

The \proglang{R} software package \pkg{margins} \citep{lenth2018package} generates marginal effects. To illustrate the distinction between effects and marginal effects, we use the simulated data described in \nameref{S1_Appendix}. 
%
\begin{figure}
\centering
\includegraphics{cubic_varpred_margins_pred.ggp.pdf}
\caption{{\bf Effect and marginal effect plots.} \NEW{A:  The effect plot. B: The marginal effect (ME) at representative values of the focal predictor \code{age} together with the $95\%$ confidence intervals. For simple linear models with no interactions, the ME should be constant across all the values of the focal predictor; and it would correspond to the slope of the central estimate of central estimate. However, in this example, the focal predictor is a cubic polynomial, and so, the effect is not constant but a function of the partial derivatives of the corresponding model variables.}}
\label{fig:qoi_age_pred_plot}
\end{figure}
%

@@

To illustrate the distinction between predictions and effects in MPVs, we simulated a multivariate model described in \nameref{S1_Appendix}. We also used the same model to illustrate how predictions and effects differ if MPVs are in the focal or non-focal predictor. In particular, we generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal predictor (a cubic polynomial focal input variable); and 2) one with \code{Wealthindex} as the focal predictor (linear focal input variable). In both cases the, we used center point as the anchor for the effects, as shown in \FREF{pred_cubic_plots}.

\begin{figure}
\begin{center}
\includegraphics{cubic_predictors_preds.ggp.pdf}
\end{center}
\caption{{\bf Prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a MPV cubic polynomial. B: The focal input variable is linear, with MPV non-focal predictor. The central estimates (dashed central curves or lines) are the same for predictions and effects, in each case (A and B). However, the effects are narrower than the predictions. The default anchor for generating effects is the center point. For linear (or simple) focal input variables, the center point is a point within the range of focal values; and therefore the effect curves intersect at that that point -- B. In the case of MPVs such as cubic polynomials, the center point is not a point but a combination of all the model variables associated with the cubic polynomial, consequently, curves will not necessarily cross at a point -- A. The horizontal black and red dashed lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively.}
\label{fig:pred_cubic_plots}
\end{figure}

In the absence of, complex interactions, MPVs, or transformations on the focal input variable, in simple OLS models we would expect the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and intersect with the trend lines at the center point, for the effect plot. On the other hand, if the focal input variable is characterized by complex interactions, MPVs, or any other form of transformations, the center point is not necessarily a point in the range of focal input values. In this case, the effects will be narrower than the predictions plots but will not necessarily intersect.

@@

Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal input variable(s), random effects or complex interactions. To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that age has a very small effect size in comparison to wealth index, and compared their effects on the estimated probability of improved water quality as shown in \FREF{pred_bin_plots}.

\begin{figure}
\begin{center}
\includegraphics{glm_two_predictor_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and whole-sample-based central estimates.} The \emph{truth} is based on the true simulation parameters. If the focal predictor has a small effect size, mean-based and whole-sample-based approaches produce different estimates -- A. In the case of strong effect size, however, the two approaches produce very close estimates -- B. In both cases, the whole-sample-based estimates are closer to the truth. The difference in mean-based and whole-sample-based is due to the bias induced by the non-focal predictor and the nonlinear link function in the logistic model. Mean-based approach suffers from nonlinear averaging; and if focal predictor has small effect size, the bias is even pronounced since the effect of nonlinear averaging is driven by the non-focal predictor(s) with the strong effect. However, this is not the case if the non-focal predictor(s) have small effect since the bias they induce is equally likely to be small. Since whole-sample-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the nonlinear averaging for the reasons already mentioned. The horizontal dashed lines correspond to the respective averages. The vertical dashed lines represent the center point, and the point at which they intersect the horizontal lines represents the expected ``perfect'' estimate at the model center. The grey points are binned observations -- observed proportions of improved water quality in each bin.} 
\label{fig:pred_bin_plots}
\end{figure}

@@

In \FREF{pred_bin_prediction_effects_plots} below, we use \FREF{pred_bin_plots}A and compare the prediction and effect plots associated with the central estimates.

\begin{figure}
\begin{center}
\includegraphics{glm_two_predictor_preds_effects.ggp.pdf}
\end{center}
\caption{{\bf Bias correction for prediction and effect plots.} In each case (A and B), the central estimates for predictions and effects are the same. These patterns are  consistent with the ones in \FREF{pred_bin_plots}A compared to the truth. Further, the predictions and effects exhibit patterns and generalize in a similar way to the ones in simple linear models.
}
\label{fig:pred_bin_prediction_effects_plots}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics{mediate_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and whole-sample-based estimates for mediated (multivariate) and non-mediated (univariate) models.} A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictions are identical since there is no any other additional (non-focal predictor) sources of bias. Consequently, in both approaches, the average estimated probability is very close to the observed proportion. B: When the mediator variable is included,  $x$ has no direct effect on $z$ and as a result the predictions do not align with the observations. However, the whole-sample-based approach still closely approximates the average proportion of the simulated data. The horizontal dashed lines are the respective average estimates and observed proportions. The vertical dashed black represents the center point, and the point at which it crosses the horizontal lines represents the expected ``perfect'' estimate at the model center. The grey points are the binned observations.}
\label{fig:pred_mediated_plots}
\end{figure}

From \FREF{pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$. This is the total effect of $x$, which only tells us the influence of $x$ on $z$. By ignoring $y$ in the model, we are still able to capture the effect of $x$ and closely match the observed values using both approaches. However, if there were additional non-focal predictors, we would expect to see the differences similar to those in \FREF{pred_bin_plots}A. Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $z$ and as a result, our estimates do not necessarily match the observed binned observations (see \FREF{pred_mediated_plots}B).

@@

The argument and results we present in this paper support a greater need for a shift on focus on how to present predictions from generalized models. For example, effect plots could provide more clarity concerning the uncertainty due to the input variable of interests as opposed to the conventional way of incorporating everything.
